---
title: "EDA Project Part 2"
author: "Data 350"
date: "Due date: April 13 2026 at midnight (EST)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this portion of the project, you will improve upon your previous EDA report to build an interactive dashboard using Shiny.
For the report we will use weather data from [Open Metro](https://open-meteo.com/). Specifically the data can be scrapped from this url https://api.open-meteo.com/v1/forecast and the data is open-sourced so no api key is needed.

```{r}
library(httr)
library(jsonlite)
```

```{r}
base_url <- "https://api.open-meteo.com/v1/forecast"

washington_dc <- list(
  latitude = 38.9072,
  longitude = -77.0369
)

response <- GET(
  url = base_url,
  query = list(
    latitude = washington_dc$latitude,
    longitude = washington_dc$longitude,

    # hourly variables
    hourly = paste(
      "temperature_2m", 
      "precipitation", 
      "precipitation_probability",
      sep = ","
    ),

    # daily summary variables
    daily = paste(
      "temperature_2m_max", 
      "temperature_2m_min", 
      "precipitation_sum", 
      sep = ","
    ),

    temperature_unit = "fahrenheit",
    precipitation_unit = "inch",
    wind_speed_unit = "mph",
    timezone = "America/New_York",
    forecast_days = 7
  )
)

stop_for_status(response) # throws an error if 4xx / 5xx
body <- content(response, as = "text", encoding = "UTF-8")
data <- fromJSON(body, simplifyVector = TRUE)

daily_df <- as.data.frame(data$daily)
names(daily_df) <- c(
  "date",
  "max_temp",
  "min_temp",
  "precip_sum"
)

hourly_df <- as.data.frame(data$hourly)
names(hourly_df) <- c(
  "time",
  "temp_2m",
  "precip_in",
  "precip_prob_pct"
)
```

```{r}
# historical data
washington_dc <- list(
  latitude  = 38.9072,
  longitude = -77.0369
)

response <- GET(
  url   = "https://archive-api.open-meteo.com/v1/archive",
  query = list(
    latitude = washington_dc$latitude,
    longitude = washington_dc$longitude,
    start_date = "2024-01-01",
    end_date = "2024-12-31",
    hourly = paste("temperature_2m", "precipitation", sep = ","),
    daily = paste("temperature_2m_max", "temperature_2m_min", "precipitation_sum", sep = ","),
    temperature_unit = "fahrenheit",
    precipitation_unit = "inch",
    wind_speed_unit = "mph",
    timezone = "America/New_York"
  )
)

stop_for_status(response)
data <- fromJSON(
  content(response, as = "text", encoding = "UTF-8"),
  simplifyVector = TRUE
)

historical_daily_df <- as.data.frame(data$daily)
names(historical_daily_df) <- c("date", "max_temp", "min_temp", "precip_sum_in")

historical_hourly_df <- as.data.frame(data$hourly)
names(historical_daily_df) <- c("time", "temp_2m", "precip_in")
```

```{r}
# install packages (valentina)
library(shiny)
library(httr)
library(jsonlite)
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
```

```{r}
# scrape data (valentina)
FIRST_YEAR <- 1940L
LAST_YEAR  <- 2026L

DC <- list(latitude = 38.9072, longitude = -77.0369)

STD_MD <- c(Spring = "03-01", Summer = "06-01", Autumn = "09-01", Winter = "12-01")

season_colors <- c(Spring = "#5DB88A", Summer = "#E8924A", Autumn = "#C06030", Winter = "#5B8DC4")

fetch_archive_slice <- function(start_date, end_date) {
  resp <- GET(
    url   = "https://archive-api.open-meteo.com/v1/archive",
    query = list(
      latitude = DC$latitude,
      longitude = DC$longitude,
      start_date = format(start_date, "%Y-%m-%d"),
      end_date = format(end_date,   "%Y-%m-%d"),
      daily = paste("temperature_2m_max", "temperature_2m_min", "precipitation_sum", sep = ","),
      temperature_unit = "fahrenheit",
      precipitation_unit = "inch",
      timezone = "America/New_York"
    )
  )
  stop_for_status(resp)
  raw <- fromJSON(content(resp, as = "text", encoding = "UTF-8"), simplifyVector = TRUE)
  df <- as.data.frame(raw$daily)
  names(df) <- c("date", "max_temp", "min_temp", "precip_sum")
  df$date <- as.Date(df$date)
  df$avg_temp <- (df$max_temp + df$min_temp) / 2
  df
}

fetch_historical <- function(progress = NULL) {
  chunk_size <- 10L
  starts <- seq(FIRST_YEAR, LAST_YEAR, by = chunk_size)
  n <- length(starts)
  chunks <- vector("list", n)

  for (i in seq_along(starts)) {
    y0 <- starts[i]
    y1 <- min(y0 + chunk_size - 1L, LAST_YEAR)
    sd <- as.Date(paste0(y0, "-01-01"))
    ed <- min(as.Date(paste0(y1, "-12-31")), Sys.Date() - 1)

    if (!is.null(progress)) {
      progress$set(
        value = (i - 1) / n,
        message = paste0("Fetching ", y0, "\u2013", y1, "\u2026")
      )
    }

    chunks[[i]] <- tryCatch(
      fetch_archive_slice(sd, ed),
      error = function(e) {
        warning("Chunk ", y0, "\u2013", y1, " failed: ", conditionMessage(e))
        NULL
      }
    )
  }

  bind_rows(Filter(Negate(is.null), chunks))
}
```

```{r}
# process data (valentina)

classify_seasons <- function(df, window = 7, summer_thresh = 65, winter_thresh = 41) {
  df <- df |> arrange(date)
  
  df$roll_avg <- as.numeric(stats::filter(df$avg_temp, rep(1 / window, window), sides = 1))
  
  df <- df[!is.na(df$roll_avg), ]
  
  df$thermal <- dplyr::case_when(
    df$roll_avg >= summer_thresh ~ "Summer",
    df$roll_avg < winter_thresh ~ "Winter",
    TRUE ~ "Transition"
  )

  df$year <- lubridate::year(df$date)

  df <- df |>
    group_by(year) |>
    group_modify(function(yr, key) {
      peak_idx <- which.max(yr$roll_avg)
      yr$season <- yr$thermal
      yr$season[seq_len(peak_idx)][yr$thermal[seq_len(peak_idx)] == "Transition"] <- "Spring"
      if (peak_idx < nrow(yr)) {
        tail_idx <- seq(peak_idx + 1L, nrow(yr))
        yr$season[tail_idx][yr$thermal[tail_idx] == "Transition"] <- "Autumn"
      }
      yr
    }) |>
    ungroup()

  df
}

# determine how long a season lasts for 

season_runs <- function(df) {
  df |>
    arrange(date) |>
    mutate(grp = cumsum(season != lag(season, default = ""))) |>
    group_by(grp, season) |>
    summarise(
      year = first(lubridate::year(date)),
      start = min(date),
      end = max(date),
      duration = as.integer(end - start) + 1L,
      mean_temp = round(mean(avg_temp, na.rm = TRUE), 1),
      .groups = "drop"
    ) |>
    arrange(start)
}

# calculate statistics for each season

yearly_season_stats <- function(runs_df) {
  runs_df |>
    group_by(year, season) |>
    slice_min(start, n = 1, with_ties = FALSE) |>
    ungroup() |>
    mutate(
      std_md = STD_MD[season],
      std_date = as.Date(paste0(year, "-", std_md)),
      deviation = as.integer(start - std_date)
    ) |>
    select(year, season, start, end, duration, mean_temp, std_date, deviation)
}
```

```{r}
# build ui (valentina)

theme_app <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#e0ddd4"),
      plot.background = element_rect(fill = "#fff", color = NA),
      panel.background = element_rect(fill = "#fff", color = NA),
      plot.subtitle = element_text(color = "#888780", size = 11),
      plot.caption = element_text(color = "#888780", size = 10)
    )
}

ui <- fluidPage(
  tags$head(tags$style(HTML("
    body { font-family: 'Helvetica Neue', Helvetica, Arial, sans-serif; background: #f8f7f4; color: #2c2c2a; }
    .well { background: #fff; border: 1px solid #e0ddd4; border-radius: 10px; box-shadow: none; }
    h2 { font-weight: 500; font-size: 22px; margin-top: 6px; }
    .section-label { font-size: 12px; color: #888780; text-transform: uppercase; letter-spacing: .05em; margin: 18px 0 6px; }
  "))),

  titlePanel(div(
    h2("Does each season start on time? How long does it last?"),
    tags$p(style = "color:#888780;font-size:14px;margin-top:2px;", "Please wait a few seconds to load all of the data!")
  )),

  sidebarLayout(
    sidebarPanel(
      width = 3,

      div(class = "section-label", "Status"),
      textOutput("last_updated"),
      br(),
      actionButton("load_data", "\u21bb  Refresh data", style = "width:100%;border-radius:8px;"),

      hr(),

      div(class = "section-label", "Season definition"),
      sliderInput("summer_thresh", "Summer threshold (\u00b0F)", min = 55, max = 80, value = 65, step = 1),
      sliderInput("winter_thresh", "Winter threshold (\u00b0F)", min = 28, max = 50, value = 41, step = 1)
    ),

    mainPanel(
      width = 9,

      tabsetPanel(
        id = "tabs",

        tabPanel(
          "Temperature Timeline", 
          br(), 
          plotOutput("temp_plot"),
          br(), br(), br(), br(), br(), br(), br(), br(), br(), br(),
          br(), br(), br(), br(), br(), br(), br(), br(), br(), br(),
          br(), br(), br(), br(), br(), br(), br(), br(), br(), br(),
          br(), br(), br(), br(), br(), br(), br(), br(), br(), br(),
          br(), br(), br(), br(), br(), br(), br(), br(), br(), br(),
          br(), br(), br(), 
          div(style = "font-size:13px;color:#5f5e5a;",
            "This graph displays the approximate start of each season through calculating the average temperature of every day since 1940, and determining when a season has started using temperature thresholds that the user can change. The graph is faceted by year, to show each year's data concurrently. Within each year, there is a line graph to show daily average temperature throughout the year, as well as color-coding to show which season that day belongs in. There are also dotted lines passing through the graph to show when the meteorological start dates for each season are.",
            "The question I wanted to explore with this graph is whether or not the duration or arrival of seasons has been affected by climate change. My hypothesis is that due to climate change, we may see seasons like spring arrive earlier, or summar last longer. However, simply through taking a look at this data, it does not appear like there is any particular trend with season arrival time or season duration changing since 1940.",
            "Potential explanations behind this are natural variations behind weather affecting season start and end time - there are many factors that can affect temperature, such as cold fronts, warm fronts, or storms, and all of these confounding factors can make it extremely difficult to establish what exactly a baseline for when the seasons start is. Additionally, there may also be issues related to our data itself - since our data only spans back until 1940, there may not be sufficient data to show the long-term effects of climate change."
          )
        ),

        tabPanel(
          "Arrival Trends",
          br(),
          plotOutput("deviation_trend_plot", height = "420px"),
          br(),
          div(style = "font-size:13px;color:#5f5e5a;",
            "This graph displays how many days early (negative) or late (positive) a season arrived compared to its meteorological standard start (Mar 1 for spring, Jun 1 for sumer, Sep 1 for fall, Dec 1 for winter). It also includes the long-run trend. This visualization uses the same data as the Temperature Timeline tab, but displays it in a different manner - here, the focus is deviation from the standard start date.", 
            "The question I wanted to explore with this graph is whether or not we see a trend with seasons starting earlier or later. My hypothesis is that due to climate change, we might expect to see seasons like spring or summer start earlier, as temperatures would surpass the temperature thresholds we define for spring or summer at an earlier date. However, simply through taking a look at this data, it does not appear like there is any particular trend with season arrival dates varying since 1940.",
            "Similarly to the previous graph, potential explanations behind this are natural variations behind weather affecting season start dates - these confounders are almost impossible to control for, and with just this dataset, we have no information on various phenomena that may have affected the data, like cold fronts or warm fronts. Additionally, the issue of simply not having enough data is also applicable here - we may not see any trends because our data may not extend far back enough."
          )
        ),

        tabPanel(
          "Duration Trends",
          br(),
          plotOutput("duration_trend_plot", height = "420px"),
          br(),
          div(style = "font-size:13px;color:#5f5e5a;",
            "This graph displays how many long a season lasted compared (represented through the points and the solid line) to the long-run average (represented as the dashed line). This visualization uses the same data as the previous tabs, but displays it in a different manner - here, the focus is season duration.", 
            "The question I wanted to explore with this graph is whether or not we see a trend with seasons lating longer or shorter. My hypothesis is that due to climate change, we might expect to see seasons like summer last longer or winter last shorter, as temperatures would surpass the temperature thresholds we define for winter or summer at an earlier date and for longer. However, simply through taking a look at this data, it does not appear like there is any particular trend with season duration varying since 1940.",
            "Once again, similarly to the previous graphs, a reason behing this could be confounding variables that we cannot control for. Since weather is so unpredictable, there are other factors that could be affecting temperatures in ways that make climate change seem more subtle based on the visualizations I've created here. Additionally, the issue of data also stands, since our dataset only ranges to 1940, which may not be sufficient to show the whole scope of climate change."
          )
        )
      )
    )
  )
)

# server stuff 

server <- function(input, output, session) {

  hist_data <- reactiveVal(NULL)
  last_fetch <- reactiveVal(NULL)

  do_load <- function() {
    progress <- shiny::Progress$new(session, min = 0, max = 1)
    on.exit(progress$close())
    progress$set(message = "Connecting to Open-Meteo\u2026", value = 0)

    tryCatch({
      hist_data(fetch_historical(progress = progress))
      last_fetch(Sys.time())
      progress$set(value = 1, message = "Done.")
    }, error = function(e) {
      showNotification(paste("Error:", conditionMessage(e)), type = "error", duration = 10)
    })
  }

  # load data 
  observe({
    do_load()
  }) |> bindEvent(TRUE, once = TRUE, ignoreInit = FALSE)

  # refresh data 
  observeEvent(input$load_data, { do_load() })

  output$last_updated <- renderText({
    req(last_fetch())
    paste("Updated:", format(last_fetch(), "%H:%M:%S"))
  })

  # classify data 

  classified <- reactive({
    req(hist_data())
    classify_seasons(hist_data(), window = 7, summer_thresh = input$summer_thresh, winter_thresh = input$winter_thresh)
  })

  runs <- reactive({
    req(classified())
    season_runs(classified())
  })

  yr_stats <- reactive({
    req(runs())
    yearly_season_stats(runs())
  })

  # tab 1: temperature trends 

  output$temp_plot <- renderPlot({
    req(classified())
    df <- classified()

    df$doy <- as.integer(format(df$date, "%j"))

    band_df <- df |>
      group_by(year, doy) |>
      summarise(season = season[1], .groups = "drop") |>
      arrange(year, doy) |>
      group_by(year) |>
      mutate(grp = cumsum(season != lag(season, default = ""))) |>
      group_by(year, grp, season) |>
      summarise(xmin = min(doy) - 0.5, xmax = max(doy) + 0.5, .groups = "drop")

    std_doy <- data.frame(
      season = names(STD_MD),
      doy = as.integer(format(as.Date(paste0("2001-", STD_MD)), "%j"))
    )

    month_breaks <- as.integer(format(as.Date(paste0("2001-", sprintf("%02d", 1:12), "-01")), "%j"))
    month_labels <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")

    n_years <- length(unique(df$year))

    ggplot(df, aes(x = doy, y = avg_temp)) +
      geom_rect(data = band_df, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = season), inherit.aes = FALSE, alpha = 0.18) +
      scale_fill_manual(values = season_colors, name = NULL) +
      geom_line(color = "#2c2c2a", linewidth = 0.4) +
      geom_vline(data = std_doy, aes(xintercept = doy), color = "#888780", linetype = "dashed", linewidth = 0.4) +
      facet_wrap(~ year, ncol = 1, strip.position = "left") +
      scale_x_continuous(breaks = month_breaks, labels = month_labels, expand = c(0, 0)) +
      scale_y_continuous(breaks = c(20, 50, 80)) +
      labs(x = "Month", y = "Year & Temperature (F)", title = "Daily Average Temperatures by Year (1940-2026)") +
      theme_app() +
      theme(
        legend.position = "bottom",
        strip.text.y.left = element_text(size = 7, angle = 0, hjust = 1, margin = margin(r = 4)),
        panel.spacing.y = unit(1, "pt"),
        axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 8),
        panel.grid.major = element_blank()
      )
  }, height = function() 18 * length(unique(classified()$year)))

  # tab 2: arrival trends

  output$deviation_trend_plot <- renderPlot({
    req(yr_stats())
    df <- yr_stats()[!is.na(yr_stats()$deviation), ]

    ggplot(df, aes(x = year, y = deviation, color = season)) +
      geom_hline(yintercept = 0, color = "#b4b2a9", linewidth = 0.5, linetype = "dashed") +
      geom_point(alpha = 0.55, size = 1.8) +
      geom_smooth(method = "loess", span = 0.35, se = TRUE, linewidth = 1, alpha = 0.15) +
      facet_wrap(~ season, ncol = 2, scales = "free_y") +
      scale_color_manual(values = season_colors, guide = "none") +
      scale_x_continuous(breaks = pretty_breaks(6)) +
      labs(
        x = "Year",
        y = "Deviation From Standard Start (Days)",
        title = "Season Arrival Since 1940"
      ) +
      theme_app() +
      theme(strip.text = element_text(face = "bold", size = 12), panel.spacing = unit(1.2, "lines"))
  })

  # tab 3: duration trends 

  output$duration_trend_plot <- renderPlot({
    req(yr_stats())
    df <- yr_stats()[!is.na(yr_stats()$duration), ]

    avg_dur <- df |>
      group_by(season) |>
      summarise(mean_dur = mean(duration, na.rm = TRUE), .groups = "drop")

    ggplot(df, aes(x = year, y = duration, color = season)) +
      geom_line(alpha = 0.45, linewidth = 0.6) +
      geom_point(alpha = 0.55, size = 1.4) +
      geom_smooth(method = "loess", span = 0.35, se = TRUE, linewidth = 1, alpha = 0.12) +
      geom_hline(data = avg_dur, aes(yintercept = mean_dur, color = season), linetype = "dashed", linewidth = 0.6, alpha = 0.7) +
      facet_wrap(~ season, ncol = 2, scales = "free_y") +
      scale_color_manual(values = season_colors, guide = "none") +
      scale_x_continuous(breaks = pretty_breaks(6)) +
      labs(
        x = "Year",
        y = "Duration (Days)",
        title = "Season Duration Since 1940"
      ) +
      theme_app() +
      theme(strip.text = element_text(face = "bold", size = 12), panel.spacing = unit(1.2, "lines"))
  })
}

shinyApp(ui, server)
```

## Preparation

1. Study the feedback that your peers gave you to see if they make sense. If so, figure out how to incorporate it into the project. Note that each group member may have received comments from different peers, so gather them.  I will work on returning feedback to each group asap.

2.The EDA projects should have an overall theme, i.e., how do you describe the goal of of your project in 1 or 2 sentences? 

## Instruction for your Shiny app

1. The app must have an well-organized interface
2. For the visualization:

    - Each question must have at least on interactive figure
    - Each questiot must use a different plot type (okay if say quesiton is `geom_point() + geom_smooth()` and question is `geom_boxplot + geom_smooth()`)
    - Each plot must include at least 4 variables
    - Please incorporate statistics (correlation and standard error), regression models and histogram distributions into the project.

## Submission

Each group member should please turn in a `RMD` file for the shiny app and a `HTML` file of the compiled application. Also please include a seperate `RMD` file for the post production.

## Grading structure

The precise rubric will be published via ELMS. In addition to the criteria from the previous projects, you will be graded on the functionality of your Shiny app and the quality of your questions/report. 

## Group Structure 

> **Optional but recommended:** switching your role from the previous project for a more well-rounded experience.

> **Caution:** Please let me know of communication issues as soon as possible. You are not expected to pick up substantial slacks from your group members. 

| Role | Responsibilities |
|------|----------------|
| **Project Leader/Manager** | - Guides the overall direction of the EDA project <br> - Organizes workflow, deadlines, communication (with team members and the instructor), and file submission  |
|||
| **Data/Stat Lead** | - Performs data cleaning, transformations, and statistical analysis <br>- Provides technical support to others |
|||
| **Visualization/Communication Lead** | - Designs clear, well-labeled, and statistically informed figures <br>- Standardizes figures/texts and integrate them into report<br>|
|||
| **All Members** | - Each member is responsible for $\geq 1$ question/figure <br> - Each member is responsible for writing a group member review, to be submitted separately |

## Post-production

Each member is required to submit a report including an individual reflection and a group member evaluation. Both will be available only to the instructor/grader. Each report must include the following key information:

- Your roles and your plots (1 paragraph)
- Summary of your work (1 paragraph)
- Evaluation of each of your team member's work, i.e., what aspects are done well and what could be improved for each member. (1 paragraph for each member)