
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
## Overview

In this portion of the project, you will improve upon your previous EDA report to build an interactive dashboard using Shiny.
For the report we will use weather data from [Open Metro](https://open-meteo.com/). Specifically the data can be scrapped from this url https://api.open-meteo.com/v1/forecast and the data is open-sourced so no api key is needed.

```{r libraries}
library(httr)
library(jsonlite)
library(dplyr)
library(lubridate)
library(ggplot2)
library(zoo)
library(shiny)
```


```{r fetch-forecast}
response <- GET(
  url = "https://api.open-meteo.com/v1/forecast",
  query = list(
    latitude           = 38.9072,
    longitude          = -77.0369,
    hourly             = paste("temperature_2m", "precipitation",
                               "precipitation_probability", sep = ","),
    daily              = paste("temperature_2m_max", "temperature_2m_min",
                               "precipitation_sum", sep = ","),
    temperature_unit   = "fahrenheit",
    precipitation_unit = "inch",
    wind_speed_unit    = "mph",
    timezone           = "America/New_York",
    forecast_days      = 7
  )
)
stop_for_status(response)
data <- fromJSON(content(response, as = "text", encoding = "UTF-8"),
                 simplifyVector = TRUE)

daily_df <- as.data.frame(data$daily)
names(daily_df) <- c("date", "max_temp", "min_temp", "precip_sum")

hourly_df <- as.data.frame(data$hourly)
names(hourly_df) <- c("time", "temp_2m", "precip_in", "precip_prob_pct")
```


```{r fetch-historical}
fetch_year <- function(year) {
  resp <- GET(
    "https://archive-api.open-meteo.com/v1/archive",
    query = list(
      latitude           = 38.9072,
      longitude          = -77.0369,
      start_date         = paste0(year, "-01-01"),
      end_date           = paste0(year, "-12-31"),
      daily              = paste("temperature_2m_max", "temperature_2m_min",
                                 "precipitation_sum", sep = ","),
      hourly             = paste("temperature_2m", "precipitation", sep = ","),
      temperature_unit   = "fahrenheit",
      precipitation_unit = "inch",
      timezone           = "America/New_York"
    )
  )
  stop_for_status(resp)
  d <- fromJSON(content(resp, as = "text", encoding = "UTF-8"),
                simplifyVector = TRUE)

  daily <- as.data.frame(d$daily)
  names(daily) <- c("date", "max_temp", "min_temp", "precip_sum_in")

  hourly <- as.data.frame(d$hourly)
  names(hourly) <- c("time", "temp_2m", "precip_in")

  list(daily = daily, hourly = hourly)
}

years <- seq(1950, 2024, by = 5)

all_daily  <- list()
all_hourly <- list()

for (yr in years) {
  message("Fetching ", yr, "...")
  result <- fetch_year(yr)
  all_daily[[as.character(yr)]]  <- result$daily
  all_hourly[[as.character(yr)]] <- result$hourly
  Sys.sleep(0.5)
}

historical_daily_df  <- bind_rows(all_daily)
historical_hourly_df <- bind_rows(all_hourly)
```


```{r prep}
# Daily data
historical_daily_df$date      <- as.Date(historical_daily_df$date)
historical_daily_df$range_f   <- historical_daily_df$max_temp - historical_daily_df$min_temp
historical_daily_df$month     <- month(historical_daily_df$date, label = TRUE, abbr = TRUE)
historical_daily_df$month_num <- month(historical_daily_df$date)
month_lookup <- setNames(1:12, month.abb)

historical_daily_df$year      <- year(historical_daily_df$date)

# Hourly data — explicit format to handle ISO 8601 timestamps from the API
historical_hourly_df$time <- as.POSIXct(
  historical_hourly_df$time,
  format = "%Y-%m-%dT%H:%M",
  tz     = "America/New_York"
)
historical_hourly_df$hour <- as.integer(hour(historical_hourly_df$time))
historical_hourly_df$date <- as.Date(historical_hourly_df$time,
                                      tz = "America/New_York")
historical_hourly_df$year  <- year(historical_hourly_df$date)
historical_hourly_df$month <- month(historical_hourly_df$date, 
                                     label = TRUE, abbr = TRUE)
historical_hourly_df$month_num <- month(historical_hourly_df$date)

# Annual summary — one row per sampled year, used for the trend tab
annual_summary <- historical_daily_df %>%
  group_by(year) %>%
  summarise(
    avg_range    = mean(range_f, na.rm = TRUE),
    sd_range     = sd(range_f, na.rm = TRUE),
    extreme_days = sum(range_f > 30, na.rm = TRUE),
    avg_max      = mean(max_temp, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r shiny-app}
# ── Helpers ──────────────────────────────────────────────────────────────────

fetch_forecast <- function() {
  resp <- GET(
    "https://api.open-meteo.com/v1/forecast",
    query = list(
      latitude           = 38.9072,
      longitude          = -77.0369,
      daily              = paste("temperature_2m_max", "temperature_2m_min",
                                 "precipitation_sum", sep = ","),
      temperature_unit   = "fahrenheit",
      precipitation_unit = "inch",
      timezone           = "America/New_York",
      forecast_days      = 7
    )
  )
  stop_for_status(resp)
  d <- fromJSON(content(resp, as = "text", encoding = "UTF-8"),
                simplifyVector = TRUE)
  daily <- as.data.frame(d$daily)
  names(daily) <- c("date", "max_temp", "min_temp", "precip_sum_in")
  daily$date    <- as.Date(daily$date)
  daily$range_f <- daily$max_temp - daily$min_temp
  daily$month   <- month(daily$date, label = TRUE, abbr = TRUE)
  daily$year    <- year(daily$date)
  daily
}

valueBoxUI <- function(id, label) {
  column(4,
    wellPanel(
      style = "text-align:center; padding:10px",
      h4(textOutput(id), style = "margin:0"),
      p(label, style = "margin:0; color:grey; font-size:12px")
    )
  )
}

# ── UI ────────────────────────────────────────────────────────────────────────

ui <- fluidPage(
  titlePanel("Has Climate Change Made DC's Weather More Extreme?"),
  sidebarLayout(
    sidebarPanel(
      width = 3,
      h4("Controls"),
      sliderInput("year_range", "Year range",
            min   = 1950,
            max   = 2024,
            value = c(1950, 2024),
            step  = 5,
            sep   = ""),
      sliderInput("month_range", "Month range",
            min = 1, max = 12,
            value = c(1, 12),
            step = 1,
            sep = ""),
      actionButton("fetch_btn", "Fetch 7-day forecast",
                   class = "btn-primary", width = "100%"),
      hr(),
      h5("How to read this"),
      p("Daily range = max − min temp on a given day. A higher value means
        a more extreme swing between the coldest and hottest parts of the day.",
        style = "font-size:12px; color:grey")
    ),
    mainPanel(
      width = 9,
      tabsetPanel(
        tabPanel("Monthly spread",
          br(),
          plotOutput("plot_monthly_box", height = "400px")
        ),
        tabPanel("IQR comparison",
          br(),
          plotOutput("plot_iqr", height = "400px"),
          br(),
          uiOutput("iqr_stats"),
          )
      )
    )
  )
)

# ── Server ────────────────────────────────────────────────────────────────────

server <- function(input, output, session) {

daily_data_raw <- reactive({
  historical_daily_df %>%
    filter(
      year      >= input$year_range[1],
      year      <= input$year_range[2],
      month_num >= input$month_range[1],
      month_num <= input$month_range[2]
    )
})

hourly_data_raw <- reactive({
  historical_hourly_df %>%
    filter(
      year      >= input$year_range[1],
      year      <= input$year_range[2],
      month_num >= input$month_range[1],
      month_num <= input$month_range[2]
    )
})

# Wait 500ms after last input change before recomputing
daily_data  <- debounce(daily_data_raw,  500)
hourly_data <- debounce(hourly_data_raw, 500)

  annual_data <- reactive({
    annual_summary %>%
      filter(year >= input$year_range[1],
             year <= input$year_range[2])
  })

  # isolate() means the forecast won't re-fetch every time other inputs change
  forecast_data <- eventReactive(input$fetch_btn, {
    tryCatch(fetch_forecast(), error = function(e) NULL)
  }, ignoreNULL = FALSE)



  # ── Tab 3: monthly boxplots ───────────────────────────────────────────────
  output$plot_monthly_box <- renderPlot({
    req(nrow(daily_data()) > 0)
    ggplot(daily_data(), aes(x = factor(year), y = range_f)) +
      geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 1) +
      labs(
        title    = "Distribution of daily temperature range by year",
        x = NULL, y = "°F range"
      ) +
      theme_minimal(base_size = 13)
  })
  
  
  # ── Tab: IQR comparison ───────────────────────────────────────────────────
output$plot_iqr <- renderPlot({
  req(nrow(daily_data()) > 0)

  daily_data() %>%
    group_by(year) %>%
    summarise(
      iqr        = IQR(range_f, na.rm = TRUE),
      q25        = quantile(range_f, 0.25, na.rm = TRUE),
      q75        = quantile(range_f, 0.75, na.rm = TRUE),
      median_range = median(range_f, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    ggplot(aes(x = year)) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "steelblue", alpha = 0.3) +
    geom_line(aes(y = median_range), colour = "steelblue", linewidth = 1) +
    geom_point(aes(y = iqr), colour = "tomato", size = 2.5) +
    geom_smooth(aes(y = iqr), method = "lm", se = FALSE,
                colour = "tomato", linetype = "dashed", linewidth = 0.9) +
    labs(
      title    = "IQR of daily temperature range by year",
      subtitle = "Red dots = IQR per year with trend line. Blue ribbon = Q25–Q75 band. Blue line = median.",
      x = "Year", y = "°F"
    ) +
    theme_minimal(base_size = 13)
})

output$iqr_stats <- renderUI({
  df <- daily_data() %>%
    group_by(year) %>%
    summarise(iqr = IQR(range_f, na.rm = TRUE), .groups = "drop")

  model <- lm(iqr ~ year, data = df)
  s     <- summary(model)
  coef  <- s$coefficients
  slope <- coef[2, 1]
  pval  <- coef[2, 4]
  r2    <- s$r.squared

  tagList(
    p(strong("Slope (per year):"), sprintf("%.4f °F", slope)),
    p(strong("p-value:"), sprintf("%.4f", pval),
      if (pval < 0.05) " — statistically significant" else " — not statistically significant"),
    p(strong("R²:"), sprintf("%.4f", r2)),
    p(strong(paste("IQR in", min(df$year), ":")), sprintf("%.2f °F", df$iqr[df$year == min(df$year)])),
    p(strong(paste("IQR in", max(df$year), ":")), sprintf("%.2f °F", df$iqr[df$year == max(df$year)])),
    p(strong("Total change:"), sprintf("%.2f °F over %d years",
      df$iqr[df$year == max(df$year)] - df$iqr[df$year == min(df$year)],
      max(df$year) - min(df$year)))
  )
})

  output$vb_avg <- renderText({
    sprintf("%.1f °F", mean(daily_data()$range_f, na.rm = TRUE))
  })
  output$vb_max <- renderText({
    sprintf("%.1f °F", max(daily_data()$range_f, na.rm = TRUE))
  })
  output$vb_volatile <- renderText({
    daily_data() %>%
      group_by(month) %>%
      summarise(sd_range = sd(range_f, na.rm = TRUE), .groups = "drop") %>%
      slice_max(sd_range, n = 1) %>%
      pull(month) %>%
      as.character()
  })
  

}

shinyApp(ui, server)
```


## Preparation

1. Study the feedback that your peers gave you to see if they make sense. If so, figure out how to incorporate it into the project. Note that each group member may have received comments from different peers, so gather them.  I will work on returning feedback to each group asap.

2.The EDA projects should have an overall theme, i.e., how do you describe the goal of of your project in 1 or 2 sentences? 


## Instruction for your Shiny app


1. The app must have an well-organized interface
2. For the visualization:

@@ -39,7 +134,6 @@ For the report we will use weather data from [Open Metro](https://open-meteo.com

Each group member should please turn in a `RMD` file for the shiny app and a `HTML` file of the compiled application. Also please include a seperate `RMD` file for the post production.


## Grading structure

The precise rubric will be published via ELMS. In addition to the criteria from the previous projects, you will be graded on the functionality of your Shiny app and the quality of your questions/report. 
@@ -50,7 +144,6 @@ The precise rubric will be published via ELMS. In addition to the criteria from

> **Caution:** Please let me know of communication issues as soon as possible. You are not expected to pick up substantial slacks from your group members. 


| Role | Responsibilities |
|------|----------------|
| **Project Leader/Manager** | - Guides the overall direction of the EDA project <br> - Organizes workflow, deadlines, communication (with team members and the instructor), and file submission  |
@@ -61,18 +154,10 @@ The precise rubric will be published via ELMS. In addition to the criteria from
|||
| **All Members** | - Each member is responsible for $\geq 1$ question/figure <br> - Each member is responsible for writing a group member review, to be submitted separately |




## Post-production

Each member is required to submit a report including an individual reflection and a group member evaluation. Both will be available only to the instructor/grader. Each report must include the following key information:

- Your roles and your plots (1 paragraph)
- Summary of your work (1 paragraph)
- Evaluation of each of your team member's work, i.e., what aspects are done well and what could be improved for each member. (1 paragraph for each member)





- Evaluation of each of your team member's work, i.e., what aspects are done well and what could be improved for each member. (1 paragraph for each member)
